Natural Selection Across Three Generations of Americans

### 2024-02-22 response-to-reviewers-revision-1.qmd

More specifically, we're interested in the pattern of selection effects
across different PGIs, and how it correlates with the PGI's relationship
to income. So we want to include PGIs with small selection effects,
which we predict to have a low absolute correlation to income. Given the
sample size, these PGIs are very likely to have insignificant individual
selection effects, but we shouldn't ignore them in the cross-PGI
analyses.

When we examine relationships across scores, we indeed test for
significance, always by bootstrapping the sample.

*The authors state that the correlation between the PGI and fertility
quantifies natural selection. However, to my knowledge it is the
covariance following the Robert-Price Identity: Morrissey, M. B., L. E.
B. Kruuk, and Alastair J. Wilson. "The danger of applying the breeder's
equation in observational studies of natural populations." Journal of
evolutionary biology 23.11 (2010): 2277-2288.*

We quantify natural selection with the regression coefficient of the PGI
on relative lifetime reproductive success. At times in the original
manuscript this was wrongly referred to as a correlation but thus has
now been updated. The polygenic scores are standardized to have a
standard deviation of 1. Under a univariate model, the regression
coefficient is the covariance divided by the variance in the independent
variable. The variance is standardized to 1, so the regression
coefficient is equal to the covariance and thus the two are
interchangeable. A brief discussion of the issue is now given in the
Data section.

In the previous manuscript we used a multivariate model, controlling for
genetic principal components. We now no longer control for the PCs. Now
we residualise the polygenic scores on the PCs before standardizing the
polygenic scores. This ensures our regression estimates are exactly the
covariance between the polygenic scores and RLRS.

@@

TE TODO: Tobias, can you check the relevant para? I guess that if both
PGI and RLRS are normalized then the covariance is the correlation; but
actually isn't the correct thing covariance with the not-normalized
RLRS?

*The authors might want to discuss the expected selection in terms of
relevance. They quantify it but they do not discuss these
quantifications. This is very important shaping our expectation of
ongoing natural selection. It might also be very interesting to know
more about whether those trends follow the secular trends in
phenotypes.*

We have now elaborated on the error corrected results and added a final
paragraph to the Discussion section on this. We estimate how selection
against intelligence will affect the phenotype, finding the result is
not especially large. In particular we discuss PGIs for health
phenotypes, as a corrective to the existing literature which is perhaps
a bit over-focused on cognition and IQ.

We briefly mention that genetic change in traits is often opposite to
the phenotypic change. This highlights that environmental change is
likely to be of greater importance for phenotypic trends then genetic
change.

### 2024-02-22 hrs-selection.qmd
Added EA example Interactions (Table unfinished). Finished up all other todos

@@

In @hugh2022human we could simply look for significant differences between coefficients on RLRS among the groups, for each polygenic score. That will not work here because of the smaller sample size. At the same time, scores are not independent and not a sample from a population. To get round this, we test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group. Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education. So, we run regressions of the form

@@

Figure @fig-deltas plots values of $\delta$ for each interaction term, without bootstrapping. The majority of $\delta$s are positive for age at first birth, income and marriage but not for education. In other words, for younger parents, poorer participants, and unmarried participants, most PGS have larger effects on RLRS, but this is not true for less educated participants.

@@

```{r}
#| label: Top PGS
#| eval: false

# Is it ok to residualise PGS on PCS vs control for them?
# Frisch-Waugh-Lovell theorem: it's the same thing... But we need PGS standardized to sd of 1
# if the regression beta is to be the covariance for the Price Equation
# When you residualise on PCs *and then restandardize* then the beta *is different* than just controlling for PCs. The difference is marginal (~1% lower in the below example). Furthermore, controlling for PCs seems to have been the standard practice in Beauchamp and in Kong et al.

# I'm not sure whether we can residualise on PCs then use the error correction because PGS true, will correlate differently with the PCs. Then again I'm not sure that the correction would be right for twin H2 since the PCs will surely only correlate with the common SNPs? I've consulted the Becker et al 2021 supplementary and it doesn't give advice on the use of PCs. I'd have to think very carefully about how to do this.

# Either we justify our approach as reasonable. Or we residualise on the PCs and go with it.
# I'm in favour of the latter, in part because explaining why the effect sizes switch sign in the error correction is awkward.

reg_simple <- lm(rlrs ~ EA ,rand$variables)
reg_pc <- lm(rlrs ~ EA + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
# beta = -0.027651
reg_res_EA <- lm(`scale(EA)` ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)

reg_res_rlrs <- lm(rlrs ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)

reg_res_noscale <- lm(reg_res_rlrs$residuals ~ reg_res_EA$residuals, 
                               reg_pc$model)
# beta = -0.027651
reg2_res <- lm(reg_res_rlrs$residuals ~ scale(reg_res_EA$residuals),
                       reg_pc$model)
# beta = -0.027358

reg2_res_pgs_only <- lm(rlrs ~ scale(reg_res_EA$residuals),
                       reg_pc$model)
# beta = -0.027358

# Below I get the estimate of the interactions and their p values. 
# EA is only significant for income at p = 0.0106
# There are always traits significant after bonferroni. But usually it's autism
# Given DHJ's improvement to manuscript, I don't think we have anything more convincing to give by focusing on the top PGS.

  tidy_pgs_x_ed[[eth]] <-
    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_ed[[eth]]) <- pgs
  tidy_pgs_x_ed[[eth]] <-
    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
    filter(grepl(':', term)) |>
    mutate(edu = str_remove(term, ".*:"))

  tidy_pgs_x_inc[[eth]] <-
    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_inc[[eth]]) <- pgs
  tidy_pgs_x_inc[[eth]] <-
    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
    filter(grepl(':', term)) |>
    mutate(income_med = str_remove(term, ".*:income_med"))

  tidy_pgs_x_mar[[eth]] <-
    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_mar[[eth]]) <- pgs
  tidy_pgs_x_mar[[eth]] <-
    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
    filter(grepl(':', term)) |>
    mutate(
      married = str_remove(term, ".*:married")
    )

  tidy_pgs_x_agefb[[eth]] <-
    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
    setNames(pgs) |> 
    list_rbind(names_to = "pgs") |>
    filter(grepl(':', term)) |>
    mutate(
      agefb = str_remove(term, ".*:agefb")
    ) 
}

@@

@@

<!--# Put in some discussion of the fact the most robust findings seem show opposite change in the phenotype during the 20th century. Negative Flynn Effect... Environmental variation across time as surely more important.-->

>>>>

```{r}
#| label: Top PGS
#| eval: true

  tidy_pgs_x_ed[[eth]] <-
    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_ed[[eth]]) <- pgs
  tidy_pgs_x_ed[[eth]] <-
    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
    filter(pgs == "EA") |>
    mutate(edu = str_remove(term, ".*:"))

  tidy_pgs_x_inc[[eth]] <-
    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_inc[[eth]]) <- pgs
  tidy_pgs_x_inc[[eth]] <-
    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
    filter(pgs == "EA") |>
    mutate(income_med = str_remove(term, ".*:income_med"))

  tidy_pgs_x_mar[[eth]] <-
    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_mar[[eth]]) <- pgs
  tidy_pgs_x_mar[[eth]] <-
    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
    filter(pgs == "EA") |>
    mutate(
      married = str_remove(term, ".*:married")
    )

  tidy_pgs_x_agefb[[eth]] <-
    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
    setNames(pgs) |> 
    list_rbind(names_to = "pgs") |>
    filter(pgs == "EA") |>
    mutate(
      agefb = str_remove(term, ".*:agefb")
    ) 
}

eff_ea_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"estimate"]
se_ea_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"std.error"]
p_ea_ed <-  ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))
eff_ea_int_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"estimate"]
se_ea_int_ed <- tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"std.error"]
p_ea_int_ed <-  ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_ed[["white"]][tidy_pgs_x_ed[["white"]]$term == 'EA:edu13-17' ,"p.value"] < 0.05, "*", "")))

eff_ea_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"estimate"]
se_ea_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"std.error"]
p_ea_inc <-  ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))

eff_ea_int_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"estimate"]
se_ea_int_inc <- tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"std.error"]
p_ea_int_inc <-  ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_inc[["white"]][tidy_pgs_x_inc[["white"]]$term == 'EA:income_medHigh' ,"p.value"] < 0.05, "*", "")))

eff_ea_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"estimate"]
se_ea_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"std.error"]
p_ea_mar <-  ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))

eff_ea_int_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"estimate"]
se_ea_int_mar <- tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"std.error"]
p_ea_int_mar <-  ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_mar[["white"]][tidy_pgs_x_mar[["white"]]$term == 'EA:marriedMarried' ,"p.value"] < 0.05, "*", "")))

eff_ea_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"estimate"]
se_ea_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"std.error"]
p_ea_agefb <-  ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA' ,"p.value"] < 0.05, "*", "")))

eff_ea_int_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"estimate"]
se_ea_int_agefb <- tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"std.error"]
p_ea_int_agefb <-  ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.001, "***",
            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.01, "**",
            ifelse(tidy_pgs_x_agefb[["white"]][tidy_pgs_x_agefb[["white"]]$term == 'EA:agefbHigh' ,"p.value"] < 0.05, "*", "")))

@@

The EA polygenic score captures a key element of human capital and was trained on a GWAS sample size of one million, making it a sensible PGS with which to test the economic theory. In TABLE XXX we report the main effect of the EA polygenic score and its interaction effect on RLRS, for each of the four putative moderators. Only for the case of income is the interaction effect significant at p \< 0.05. We might wonder whether the weak effects are because of insufficient power or because any real interaction is small in the sample. In @hugh2022human the interaction effects were large, often as big as the main effects. Yet in the HRS, the point estimates of the interactions appear generally smaller, suggesting the nulls may not just be a consequence of lower power.

We should attain more statistical power if we utilise information from other polygenic scores at the same time. However, some caution is necessary. Owing to the fact that the results for each PGS are not independent we should expect an elevated false positive rate. To perform, we test the null hypothesis that there are no differences in coefficients between social groups for any PGS, against the alternative that for some PGS, coefficients are larger for the disadvantaged group. Here, "larger" means in the direction predicted by the score's correlation with education: more positive for scores which are negatively correlated with education, and more negative for scores which are positively correlated with education. So, we run regressions of the form

@@

@fig-deltas plots values of $\delta$ for each interaction term, without bootstrapping. The majority of $\delta$s are positive for age at first birth, income and marriage but not for education. In other words, for younger parents, poorer participants, and unmarried participants, most PGS have larger effects on RLRS, but this is not true for less educated participants.

@@

@@

To our knowledge, this is the first time error corrected estimates of natural selection have been made for non-cognitive traits. Given the focus on cognitive traits, it is interesting to see that the point estimates for many other traits exceed point estimates for educational attainment and cognitive performance. Depressive symptoms, asthma, age of first birth and self-rated health all appear to be changing faster. This suggests the effects of natural selection are not limited to socioeconomic outcomes, but are also of medical relevance.

@@

Looking at the polygenic scores changing the most, many seem to have moved in the opposite direction to how the phenotype has changed over the latter half of the twentieth century. Education, IQ, delayed child birth, abstinence from smoking and life expectancy have all increased, contrary to how we believe the genetics have changed. The main exception is that the prevalence of mental illnesses and depression have increased in recent years in America, paralleling the polygenic scores for depression and neuroticism, nevertheless it is unclear whether these changes are real or due to be increased diagnosis and reporting. The apparent paradox of phenotypes and nautral selection moving in opposite directions suggest that environmental change across history is a much greater force than that of natural selection.

<!--# Bit lazy perhaps not provide citations for phenotypic change, but the changes are close to common knowledge.-->
